---
title: "STAT 221 Homework 2"
author: "John Rapp Farnes | 405461225"
date: "2/13 2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 2.9

a)

```{r}
library(astsa)

model <- lm(soi ~ time(soi))

(sum <- summary(model))

sum$coefficients[2]
```
There is a significant trend in SOI in relation to time, where the SOI decreases ~0.007 every year.

b)
```{r}
plot(soi, main="SOI, trend in red")
lines(soi*0 + predict(model, time(soi)), type="l", col="red")

detrended <- soi - predict(model, time(soi))

plot(detrended, main="SOI detrended")
```
```{r}
library(TSA)
pg <- periodogram(detrended)

```


Two peaks, first at period:
```{r}
max_index <- which(pg$spec == max(pg$spec))
freq <- pg$freq[max_index]

1/freq
```
Or every year. Second at period:

```{r}
freq <- pg$freq[which(pg$spec[1:(max_index - 1)] == max(pg$spec[1:(max_index - 1)]))]

1/freq
```

Or every five years, which is the probable El Nino cycle.

\newpage

# 3.2

a)

By expanding "backwards" in time, we have

\begin{equation}
  \begin{split}
  x_t &= \phi x_{t-1} + w_t \\
  &= \phi (\phi x_{t-2} + w_{t-1}) + w_t \\
  &= \phi^2 x_{t-2} + \phi w_{t-1} + w_t \\
  &\;\;\vdots \notag \\
  &= \phi^k x_{t-k} + \sum_{j=0}^{k-1}\phi^j w_{t-j}
  \end{split}
\end{equation}

When $k = t$ we get

\begin{equation}
  \begin{split}
  x_t &= \phi^t x_{0} + \sum_{j=0}^{t-1}\phi^j w_{t-j} \\
  &= \phi^t w_{0} + \sum_{j=0}^{t-1}\phi^j w_{t-j} \\
  &= \sum_{j=0}^{t}\phi^j w_{t-j} \\
  \end{split}
\end{equation}

b)

We have

\begin{equation}
  \begin{split}
  E(x_t) &= E(\sum_{j=0}^{t}\phi^j w_{t-j}) \\
  &= \sum_{j=0}^{t}\phi^j E(w_{t-j}) = 0\\
  \end{split}
\end{equation}

as $E(w_i) = 0 \: \forall i$

c)

As $Var(w_i) = \sigma_w^2\: \forall i$, we have

\begin{equation}
  \begin{split}
  Var(x_t) &= Var(\sum_{j=0}^{t}\phi^j w_{t-j}) \\
  &= \sum_{j=0}^{t}(\phi^{j})^2 Var(w_{t-j}) \\
  &= \sigma_w^2 \sum_{j=0}^{t}\phi^{2j} \\
  &= \sigma_w^2 \frac{1-\phi^{2(t+1)}}{1-\phi^2}
  \end{split}
\end{equation}

by geometric series formula, as $|\phi| < 1 \implies |\phi^2| < 1$.

d)

\begin{equation}
  \begin{split}
  cov(x_{t+h}, x_t) &= E[(\sum_{j=0}^{t+h}\phi^j w_{t+h-j})(\sum_{j=0}^{t}\phi^j w_{t-j})] \\
  &= E[(w_{t+h} + \dots + \phi^h w_t + \phi^{h+1} w_{t-1}+\dots+\phi^{h+t} w_{0})(w_t + \phi w_{t-1} + \dots+\phi^t w_{0}]) \\
  &= \sigma_w^2 \sum_{j=0}^{t} \phi^{h+2j} \\
  &= \sigma_w^2 \phi^{h} \sum_{j=0}^{t} \phi^{2j} \\
  &= \phi^{h} \sigma_w^2 \frac{1-\phi^{2(t+1)}}{1-\phi^2} \\
  &= \phi^h Var(x_t)
  \end{split}
\end{equation}

e)
We have $E(x_t)=0$ constant, however $\gamma(h) = \phi^h Var(x_t) = \phi^h \sigma_w^2 \frac{1-\phi^{2(t+1)}}{1-\phi^2}$ depends on time. Hence, the series is not stationary.

f)
We have $t \to \infty \implies \gamma(h) \to \phi^h \sigma_w^2 \frac{1}{1-\phi^2}$ depends only on $h$ and not on $t$, hence it is "asymptotically stationary."

g)
As we have just proved that this process estimates an AR(1) process, a AR(1) process could be simulated by simulating $n$ i.i.d $N(0,1)$ noise values as $w_t, t=1..n$ and then calculating $x_t$ by

\begin{equation}
\begin{cases}
  x_t =  \phi x_{t-1}+w_t, t=1..n\dots \\
  x_0=w_0
\end{cases}
\end{equation}

h)
We have, with $k=t$,

\begin{equation}
  \begin{split}
  x_t &= \phi^t x_{0} + \sum_{j=0}^{t-1}\phi^j w_{t-j} \\
  &= \phi^t w_{0}/\sqrt{1-\phi ^2} + \sum_{j=0}^{t-1}\phi^j w_{t-j} \implies \\
  Var(x_t) &= \frac{(\phi^t)^2}{1-\phi ^2} \sigma_w^2 + \sigma_w^2 \frac{1-\phi^{2t}}{1-\phi^2} \\
  &= \frac{\sigma_w^2}{1-\phi ^2}(\phi^{2t}+1-\phi^{2t}) \\
  &= \frac{\sigma_w^2}{1-\phi ^2}
  \end{split}
\end{equation}

So $Var(x_t)$ is constant, and $E(x_t)$ is still constant, hence the series is stationary.


# 3.6

We have
\begin{equation}
  \begin{split}
  x_t &= -.9x_{t-2} + w_t \implies \\
  x_t + .9x_{t-2} &= w_t \implies \\
  (1 + 0.9B^2)x_t &= w_t
  \end{split}
\end{equation}

So $\phi(z) = 1+0.9z^2$, which has the roots $z_i$:

```{r}
(z <- polyroot(c(1,0,0.9)))
abs(polyroot(c(1,0,0.9)))
```



Hence $|z_i| > 1 \: \forall i$ and $z_i$ complex conjugate, as such the acf will have periodic behavior, see plot.

```{r}
plot(ARMAacf(ar = c(0, -0.9), lag.max = 10))
lines(rep(0, 11))
```


# 3.9
First we'll plot the simulated series

```{r}
set.seed(2020)
ar <- arima.sim(list(order=c(1,0,0), ar=.6), n=100)
ma <- arima.sim(list(order=c(0,0,1), ma=.9), n=100)
arma <- arima.sim(list(order=c(1,0,1), ar=.6, ma=.9), n=100)

par(mfrow=c(3, 1))
plot(ar, ylab="x", main=(expression(AR(1)~~~phi==+.6)))
plot(ma, ylab="x", main=(expression(MA(1)~~~theta==+.9)))
plot(arma, ylab="x", main=(expression(ARMA(1, 1)~~~phi==+.6~theta==+.9)))
```

Next, we'll plot and compare the sample acf with the theoretical acf.

```{r}
par(mfrow=c(2, 1))
plot(acf(ar, lag.max = 10, plot=FALSE), main=(expression(Sample~acf~AR(1)~~~phi==+.6)))
plot(ARMAacf(ar = 0.6, lag.max = 10)[-1], main=(expression(Theoretical~acf~AR(1)~~~phi==+.6)), ylab = "ACF", xlab="Lag")

par(mfrow=c(2, 1))
plot(acf(ma, lag.max = 10, plot=FALSE), main=(expression(Sample~acf~MA(1)~~~theta==+.9)))
plot(ARMAacf(ma = 0.9, lag.max = 10)[-1], main=(expression(Theoretical~acf~MA(1)~~~theta==+.9)), ylab = "ACF", xlab="Lag")

par(mfrow=c(2, 1))
plot(acf(arma, lag.max = 10, plot=FALSE), main=(expression(Sample~acf~ARMA(1, 1)~~~phi==+.6~theta==+.9)))
plot(ARMAacf(ar = 0.6, ma = 0.9, lag.max = 10)[-1], main=(expression(Theoretical~acf~ARMA(1, 1)~~~phi==+.6~theta==+.9)), ylab = "ACF", xlab="Lag")
```

The plots are consistent with the general results of Table 3.1, as the AR(1) as well as ARMA(1,1) process tails off while the MA(1) cutts of after lag 1$=q$.


```{r}
par(mfrow=c(2, 1))
plot(pacf(ar, lag.max = 10, plot=FALSE), main=(expression(Sample~pacf~AR(1)~~~phi==+.6)))
plot(ARMAacf(ar = 0.6, lag.max = 10, pacf = TRUE)[-1], main=(expression(Theoretical~pacf~AR(1)~~~phi==+.6)), ylab = "ACF", xlab="Lag")

par(mfrow=c(2, 1))
plot(pacf(ma, lag.max = 10, plot=FALSE), main=(expression(Sample~pacf~MA(1)~~~theta==+.9)))
plot(ARMAacf(ma = 0.9, lag.max = 10, pacf = TRUE)[-1], main=(expression(Theoretical~pacf~MA(1)~~~theta==+.9)), ylab = "ACF", xlab="Lag")

par(mfrow=c(2, 1))
plot(pacf(arma, lag.max = 10, plot=FALSE), main=(expression(Sample~pacf~ARMA(1, 1)~~~phi==+.6~theta==+.9)))
plot(ARMAacf(ar = 0.6, ma = 0.9, lag.max = 10, pacf = TRUE)[-1], main=(expression(Theoretical~pacf~ARMA(1, 1)~~~phi==+.6~theta==+.9)), ylab = "ACF", xlab="Lag")
```

The plots are consistent with the general results of Table 3.1, as the MA(1) as well as ARMA(1,1) process tails off while the AR(1) cutts of after lag 1$=q$.
